\mytitle

### 2016-10-05 pdocs/thesis/article-introduction.tex
tuning of article
Aki's comments on article

\section{Introduction}

The application of modern computer game techniques enables the description of complex dynamic systems 
such as military vehicles with a high level of detail while still solving the equations in real-time.
Film production and war games, in particular, is a key area that have benefited from simulation technology. 
In practice, games are often accomplished using an open-source platform such 
like ODE - \cite{ode}, \bullet\ - \cite{bullet} and Box2D - \cite{box2d}.

Computational methods used in physics engines are divided to modules that handle collision detection and 
contact description and modules that handle solution of equations in real-time. Equations need to be 
solved can further be subdivided to be associated to motion, constraints and collisions. 
Velocity-based formulation is typically used in constraint based rigid body simulation. 
Friction is typically taken into account and mechanical joints are handled by constraint equations.
Detailed description of various components can be found in e.g. \cite{erleben.thesis}.

Plasticity is not typically taken into account in gaming solutions. 
Breaking of various objects typically takes place based on collision or impulse.
Nevertheless, breaking of steel or reinforced concrete structures using this approach 
is not appropriate making a simulation to look unrealistic. Theory for handling of plasticity 
has been presented already in \cite{cg1988}. \cite{muller2004point} and \cite{muller2005meshless} 
present a method for modeling and animating of elastic and plastic objects in real-time using 
point based animation. This approach is not been widely used in simulation applications.  
On major issue is collision handling of deformable objects.

This study will introduce an approach to account plastic deformation in game applications.   
In the introduced method, the plastic deformation takes place if force or moment exceeds given 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
The approach is based on using joint motors to model plasticity. \citet[p.~90]{erleben.thesis} 
suggests similar method for modelling friction in joints. Adjacent objects are connected by motors. 
Motor power production limits are estimated based on plastic section modulus. 
Joint breaking is accounted by summing plastic deformation and comparing it to 
predefined material based limit. Elastic part of deformation is modelled by employing 
spring description which is based on modification of existing constraint in \bullet.

>>>>

% !TeX root = article.tex
\section{Introduction}

The application of modern computer game techniques enables the description of complex dynamic systems 
such as military vehicles with a high level of detail while still solving the equations of motion in real-time.
Film production and war games, in particular, is a key area that have benefited from rapidly enhanced simulation technology. 
In practice, games are often accomplished using an open-source platform such 
like ODE - \cite{ode}, \bullet\ - \cite{bullet} and Box2D - \cite{box2d}.

Computational methods used in physics engines are divided to modules that handle collision detection and 
contact description and modules that handle solution of equations of motion in real-time. Equations of motion need to be 
solved can further be subdivided to be associated to motion, constraints and collisions. 
Velocity-based formulation is typically used in constraint based rigid body simulation, \cite{erleben.thesis}. 
Friction is typically taken into account and mechanical joints are handled by constraint equations,
\cite{erleben.thesis}.

Plasticity is not typically taken into account in gaming solutions. 
Breaking of various objects often takes place based on collision or impulse.
Nevertheless, breaking of steel or reinforced concrete structures using this approach 
is not appropriate making a simulation to look realistic. Theory for handling of plasticity 
has been presented already in \cite{cg1988}. \cite{muller2004point} and \cite{muller2005meshless} 
present a method for modeling and animating of elastic and plastic objects in real-time using 
point based animation. This approach is not been widely used in computer games.  
On major issue is collision handling of deformable objects.

This study will introduce an approach to account plastic deformation in game applications.   
In the introduced method, the plastic deformation takes place if force or moment exceeds predefined 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
The approach is based on using joint motors to model plasticity. 
This study extends a method introduced by
\citet{erleben.thesis} 
%\citet[p.~90]{erleben.thesis} 
which is originally proposed for modelling friction in joints. 
In the introduced method adjacent objects are connected by motors. 
Motor power production limits are estimated based on plastic section modulus. 
Joint breaking is accounted by summing plastic deformation and comparing it to 
predefined material based limit. Elastic part of deformation is modelled by employing 
spring description which is based on modification of existing constraint in \bullet.

### 2016-10-05 pdocs/thesis/article-section-2.tex
Aki's comments on article

Velocity-based formulation is  popular within physics based game
developers and film production teams.
 \citet[p.~45]{erleben.thesis} provides reasoning and theoretical details on why 
velocity-based formulation is  popular in constraint-based rigid body simulation. 
Main reason is that collision handling can be done without additional procedures.

Background
for velocity based formulation shown here is based on  \citet[p.~45-50]{erleben.thesis}.
% pdf page 64
In following section, these formulations will be clarified by simple example using \bullet\ implementation.
Impulse $\vec{J}$
in the time interval $\Delta t $ can we written as
\begin{equation} \label{eq:impulseIntegral}
\vec{J} = \int_{0}^{\Delta t} \vec{f}_{true}(t) dt
\end{equation}
where $\vec{f}_{true}(t)$ is force.

Using Newton's second law of motion $\vec{F}=m\vec{a}$ one can solve for the velocity,
$\vec{v}^{\Delta t}$
\begin{equation} \label{eq:impulseIntegraWithNewton}
\int_{0}^{\Delta t} m \frac{d\vec{v}}{dt}dt= \int_{0}^{\Delta t} \vec{f}_{true}(t)
\end{equation}
\begin{equation} \label{eq:impulse}
m(\vec{v}^{\, \Delta t} - \vec{v}^{\, 0})=\vec{J}
\end{equation}
where superscripts denote time, i.e. ${\vec{v}}^{\Delta t}=\vec{v}(\Delta t)$.
Next position can be found
by integrating the velocity.
Updates after each step can be summarized as Equation
 \ref{eq:eomL} for locations and  Equation 
\ref{eq:eomV} for velocities.

@@

Symbols are summarized in Table \ref{tab:eom} and Figure \ref{fig:eom-contact}.
Figure \ref{fig:eom-contact} describes collision of two bodies $B_1$ and $B_2$

@@

where $\vec{r}_i$ is orientation of body i,
$\vec{p}_k $ is orientation of contact point k,
$\vec{r}_{ik} $ is vector  between center of gravity of body i and contact point k
and
$\vec{n}_{k}$ is contact normal for contact point k.
With this formulation, simplified scenario can be simulated.

@@

Friction in contacts and joint constraints can be handled in unified way by refactoring
equation \ref{eq:eomV} to get \ref{eq:eomV2},  \citet[p.~66-67]{erleben.thesis}
\begin{equation} \label{eq:eomV2}
\vec{u}^{\, t+\Delta t} = \vec{u}^{\, t}+\Delta t M^{-1}(
 J_{contact}^T \vec{\lambda}_{contact}
+ J_{joint}^T \vec{\lambda}_{joint}
+ \vec{f}_{ext})
\end{equation}

@@

Added symbols are summarized in Table
\ref{tab:eom-g} and Figure \ref{fig:eom-joint}.
Figure \ref{fig:eom-joint} shows terms needed for joint processing.

@@

Constraint processing in \bullet\ is based on ODE, \cite{ode}.
Mathematical background and detailed examples are available by \cite{ode.joints}.
Joints are also discussed in detail in  \citet[p.~60-90]{erleben.thesis}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint.

@@

where $\sigma$ is stress, $E$ is Youngs modulus and $f_y$ is yield stress.
Engineering stress and strain mean that original dimensions are used in stress calculation,
\citet[p.~108]{dowling}.
Stress-strain curve is not drawn to scale as elastic strain could not be seen as it is typically 0.001 to 0.005.

@@

Strain hardening is taken into account in this work mainly by assuming that plasticity in bending
expands, \citet[p.~672]{dowling}.
Material that starts to yield first is hardened and yielding moves slightly.
This can be seen e.g. by bending paperclip. It does not break at low angles but can take few full bends.

>>>>

Velocity-based formulation is  popular within physics based game
developers and film production teams.
%\citet[p.~45]{erleben.thesis} 
\citet{erleben.thesis} 
provides reasoning and theoretical details on why 
velocity-based formulation is  popular in constraint-based rigid body simulation. 
Main reason is that collision handling can be done without additional procedures.

Background
for velocity based formulation shown here is based on  
\citet{erleben.thesis}.
%\citet[p.~45-50]{erleben.thesis}.
% pdf page 64
In following section, these formulations will be clarified by simple example using \bullet\ implementation.

Impulse $\vec{J}$
in the time interval $\Delta t $ can we written as
\begin{equation} \label{eq:impulseIntegral}
\vec{J} = \int_{0}^{\Delta t} \vec{f}_{true}(t) dt
\end{equation}
where $\vec{f}_{true}(t)$ is force.

Newton's second law of motion $\vec{F}=m\vec{a}$ one can solve for the velocity,
$\vec{v}^{\Delta t}$ as
\begin{equation} \label{eq:impulseIntegraWithNewton}
\int_{0}^{\Delta t} m \frac{d\vec{v}}{dt}dt= \int_{0}^{\Delta t} \vec{f}_{true}(t)
\end{equation}
\begin{equation} \label{eq:impulse}
m(\vec{v}^{\, \Delta t} - \vec{v}^{\, 0})=\vec{J}
\end{equation}
where superscripts denote time, i.e. ${\vec{v}}^{\Delta t}=\vec{v}(\Delta t)$.
Next position can be found
by integrating the velocity.
Updates after each step can be summarized  for locations and  
for velocities correspondingly as follows.

@@

Symbols used in Equations \ref{eq:eomL} and \ref{eq:eomV}
are summarized in Table \ref{tab:eom} and Figure \ref{fig:eom-contact}.
Figure \ref{fig:eom-contact} describes collision of two bodies $B_1$ and $B_2$

@@

where $\vec{r}_i$ is orientation of body $i$,
$\vec{p}_k $ is orientation of contact point $k$,
$\vec{r}_{ik} $ is vector  between center of gravity of body i and contact point $k$
and
$\vec{n}_{k}$ is contact normal for contact point $k$.
By employing this formulation, simplified scenario can be simulated.

@@

Friction in contacts and joint constraints can be handled in unified way by refactoring
equation \ref{eq:eomV} as,  
\citet{erleben.thesis}
%\citet[p.~66-67]{erleben.thesis}
\begin{equation} \label{eq:eomV2}
\vec{u}^{\, t+\Delta t} = \vec{u}^{\, t}+\Delta t M^{-1}(
 J_{contact}^T \vec{\lambda}_{contact}
+ J_{joint}^T \vec{\lambda}_{joint}
+ \vec{f}_{ext})
\end{equation}

@@

Symbols used in Equation \ref{eq:eomV2} are summarized in Table
\ref{tab:eom-g} and Figure \ref{fig:eom-joint}.
Figure \ref{fig:eom-joint} shows terms needed for joint processing.

@@

Constraint processing in \bullet\ is based on ODE, \cite{ode}.
Mathematical background and detailed examples are available by \cite{ode.joints}.
Joints are also discussed in detail in  
\citet{erleben.thesis}.
%\citet[p.~60-90]{erleben.thesis}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint.

@@

where $\sigma$ is stress, $E$ is Youngs modulus and $f_y$ is yield stress.
Engineering stress and strain mean that original dimensions are used in stress calculation,
\citet{dowling}.
%\citet[p.~108]{dowling}.
Stress-strain curve is not drawn to scale as elastic strain could not be seen as it is typically 0.001 to 0.005.

@@

Strain hardening is taken into account in this work mainly by assuming that plasticity in bending
expands, 
\citet{dowling}.
%\citet[p.~672]{dowling}.
Material that starts to yield first is hardened and yielding moves slightly.
This can be seen e.g. by bending paperclip. It does not break at low angles but can take few full bends.

### 2016-10-05 pdocs/thesis/article.tex
pdocs/article
pdocs/article changes based on Aki's comments
first draft of section 2 for article
article tuning for section 2
article section 3 started
velocity based formulation basics added to article using erleben as source
more math details in article section 2
added maximum forces in article section 2
working on Aki's comments on article
skeleton for conclusion
Aki's comments on article

% Simo.Nikula@gmail.com
% based on latex template at http://www.lut.fi/web/en/library/dissertations1 
\def\mytitle{Handling of plasticity in physics engine}
\def\myisbn{978-952-265-XXX-X}
\def\mypdfisbn{978-952-265-XXX-X}
\def\myyear{2017}
\def\mymonth{May}
\def\myname{Simo Nikula}
\def\lut{Lappeenranta University of Technology}
\def\bullet{Bullet Physics}
\def\demolisher{DemolisherDemo}
\def\charpy{CharpyDemo}
\documentclass[a4paper,twoside,12pt,notitlepage,openright]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{lastpage}   % page count
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage[intoc]{nomencl} % including table of content
\usepackage{ifthen}
\usepackage[hmargin=3.0cm,vmargin=3.6cm]{geometry} % setting marginals
\usepackage{fancyhdr,extramarks}  % header ja footer manipulation
\usepackage{times}  % to change font to times
\usepackage{setspace} % for linespacingx
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{caption}
\usepackage{varwidth}
\newcommand{\nomunit}[1]{\renewcommand{\nomentryend}{\hspace*{\fill}#1}} % Inserts units on the right at symbol list
\renewcommand{\nomgroup}[1]{%
 \ifthenelse{\equal{#1}{C}}{\item[\textbf{Latin alphabet}]\item}{%
 \ifthenelse{\equal{#1}{G}}{\item\item[\textbf{Greek alphabet}]\item}{}}{%
 \ifthenelse{\equal{#1}{L}}{\item\item[\textbf{Subscripts}]\item}{}}{%
 \ifthenelse{\equal{#1}{H}}{\item\item[\textbf{Superscripts}]\item}{}}{%
 \ifthenelse{\equal{#1}{W}}{\item\item[\textbf{Abbreviations}]\item}{}}{}}

@@

\pagestyle{empty}
\include{nimio}
\include{abstract}
\addtocontents{toc}{\contentsline {section}{Abstract}{}}

\include{acknowledgments}
\addtocontents{toc}{\contentsline {section}{Acknowledgments}{}}%
\input{preface}
\cleardoublepage%
\thispagestyle{empty}%
\addtocontents{toc}{\contentsline{section}{Contents}{}}%

\tableofcontents%
\thispagestyle{empty}%
\cleardoublepage%
\pagestyle{fancy}%

\cleardoublepage%
\printnomenclature[2.0cm]%

\include{introduction}
\include{bullet-plasticity}
%\include{results}
\include{discussion}

%\def\refname{what ever}% to change References title
\addcontentsline{toc}{section}{References}%

\fancyhead[LO]{\nouppercase{\bfseries\firstleftxmark}}%
\fancyhead[RE]{\nouppercase{\bfseries\lastrightxmark}}%
\extramarks{}{References}   % this will remove the "References" to be appearing at the header at first page of bibliography
\extramarks{References}{References}
\bibliography{ref}%
\bibliographystyle{LUTapa}%

>>>>

% Simo.Nikula@gmail.com
% based on latex template at http://www.lut.fi/web/en/library/dissertations1 
\def\mytitle{Description of plasticity in physics engine}
\def\myisbn{978-952-265-XXX-X}
\def\mypdfisbn{978-952-265-XXX-X}
\def\myyear{2017}
\def\mymonth{May}
\def\myname{Simo Nikula}
\def\lut{Lappeenranta University of Technology}
\def\bullet{Bullet Physics}
\def\demolisher{DemolisherDemo}
\def\charpy{CharpyDemo}
\documentclass[a4paper,twoside,12pt,notitlepage]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{lastpage}   % page count
\usepackage{amsmath}
\usepackage{listings}
% \usepackage{mathdots}  % improved \ddots and \iidots
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage[intoc]{nomencl} % including table of content
\usepackage{ifthen}
\usepackage[hmargin=2.0cm,vmargin=1.6cm]{geometry} % setting marginals
\usepackage{fancyhdr,extramarks}  % header ja footer manipulation
\usepackage{times}  % to change font to times
\usepackage{setspace} % for linespacingx
\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\usepackage{caption}
\usepackage{varwidth}
\usepackage{enumitem} % 
\newcommand{\nomunit}[1]{\renewcommand{\nomentryend}{\hspace*{\fill}#1}} % Inserts units on the right at symbol list
\renewcommand{\nomgroup}[1]{%
 \ifthenelse{\equal{#1}{C}}{\item[\textbf{Latin alphabet}]\item}{%
 \ifthenelse{\equal{#1}{G}}{\item\item[\textbf{Greek alphabet}]\item}{}}{%
 \ifthenelse{\equal{#1}{L}}{\item\item[\textbf{Subscripts}]\item}{}}{%
 \ifthenelse{\equal{#1}{H}}{\item\item[\textbf{Superscripts}]\item}{}}{%
 \ifthenelse{\equal{#1}{W}}{\item\item[\textbf{Abbreviations}]\item}{}}{}}

@@

\pagestyle{empty}
\title{\mytitle}
\author{Simo Nikula, Aki Mikkola and Timo Björk}
\author{}
\maketitle
\include{article-introduction}
\include{article-section-2}
\include{article-section-3}
\include{article-conclusion}

\fancyhead[LO]{\nouppercase{\bfseries\firstleftxmark}}%
\fancyhead[RE]{\nouppercase{\bfseries\lastrightxmark}}%
\extramarks{}{References}   % this will remove the "References" to be appearing at the header at first page of bibliography
\extramarks{References}{References}
\bibliographystyle{LUTapa}% LUTapa
\bibliography{ref}%
